Stable individual differences in strategies within, but not between, visual search tasks

### 2019-02-14 Manuscript/manuscript.tex
shortened to try and meet length restrictions

As is common in cognitive psychology, most visual search literature has focused on how the average participant performs in the task, despite it being well known that there is a great deal of variability between one subject and the next. From Treisman's work on Feature Integration Theory \citep{treisman1980} to the latest incarnation of the Guided Search Model \citep{wolfe2015}, we have a good understanding of what makes some search targets effortless to find, while others require careful inspection. However, these theories and models have neglected the question of why some observers find visual search so much harder than others. These differences can emerge from several different sources of variation: tiredness \citep{mackworth1948}, information-processing ability, speed-accuracy trade-off, motivation, visual impairments \citep{nowakowska2016}, and search strategies \citep{boot2006}. Although their existence has previously been noted \citep{mackworth1948}, a rigorous examination of individual differences in visual search is a challenge that has not been taken up by many researchers, and questions about their importance and stability remain relatively under explored.

In the current study, we focus on one source of individual differences in visual search: strategy. Here we refer to a collection of behaviours that all observers can freely choose from when they search a display. Examples include choosing to adopt a systematic left-to-right and top-to-bottom strategy \citep{gilchrist2006}, or the type of guided search behaviour in which locations that are more likely to contain the target are prioritised \citep{wolfe2015}.

A striking example of the effect of strategy is given by \cite{boot2006}. They asked participants to monitor a cluttered display for an object changing colour or suddenly appearing. Large individual differences were found with respect to the number of saccades participants made while monitoring the stimulus, which was negatively correlated with detection performance. Proulx \citeyearpar{proulx2011} demonstrated that participants differed in how they prioritised visual features (colour or orientation) when searching through a set of objects for one with a particular colour-orientation combination. These differences in search strategy explained some of the variation in the extent to which participants were distracted by an irrelevant object of a unique size. However, participants were largely unaware of the search strategy they employed, and the majority were unable, or unwilling, to comply with instructions to employ a particular strategy.

Eye movement strategies have also been shown to be an important source of individual differences in visual search efficiency. \cite{nowakowska2017} designed a simple search paradigm to discriminate between optimal \citep{najemnik-geisler2008} and stochastic  \citep{clarke2016} search strategies. The task involved searching through arrays of line segments (Figure \ref{fig:exampleStimuli}) arranged such that those on one side of the display all had a very similar orientation (homogeneous), while those on the other side had higher variance (heterogeneous). This meant that targets appearing on the homogeneous side would be highly salient, while targets on the heterogeneous side  would be harder to find. With these stimuli, the optimal eye movement strategy is to search the heterogeneous half, as targets on the homogeneous side can be detected with peripheral vision. We will refer to this paradigm here as the Split-Half Line Segment task (SHLS). While some participants initially searched the displays near optimally, others carried out strategies counter to this, failing to even match the performance of the stochastic searcher. Furthermore, the degree to which they made saccades in line with the optimal search strategy was strongly correlated with their reaction times. A related version of this paradigm has been used in research investigating eye movement strategies in response to (simulated) hemianopia \citep{nowakowska2016,nowakowska2018}, with similar conclusions: the full spectrum of individual differences in strategy was observed, from near-optimal to the opposite of optimal. It is therefore not possible to conclude whether optimal or stochastic models better describe search without first explaining individual variability.

@@

Another example of differences in search strategy comes from the foraging literature  \citep{kristjansson2014,johannesson2016}. In this context, foraging is a visual search task in which there are multiple targets on each trial. Participants were asked to search through a set of items from four categories, with two categories classed as targets. In the conjunction condition (searching for red-horizontal and green-vertical line segments among red-vertical and green-horizontal distracters), most observers searched in runs, finding all the targets of one target category, and then switching and finding the targets in the other category. This strategy has previously been observed in animal foraging  \citep{dawkins1971}, and suggests holding one complex target template in mind at a time is a better strategy than switching templates. However, a sub-set of observers, termed `super-foragers', were able to change between search target categories with no cost to performance. This study has been run on touch displays \citep{kristjansson2014} (observers are asked to tap on as many targets as they can as quickly as possible) and a 3D computer game (in which participants navigate a squirrel through a virtual environment picking up targets with different feature combinations) \citep{prpic2018}, with super-foragers present in both versions. While test-retest reliability has not been measured explicitly for the foraging paradigm, the task was used as a measure to assess the effect of a six day long mindfulness retreat on cognitive performance \citep{hartkamp2017}. From a re-analysis of these data, we can estimate that the test-retest reliability for the mean run length is $r \approx 0.7$ for the feature condition and $r \approx 0.88$ for the conjunction search.

@@

A common theme emerging from these studies is the observation that individual strategies vary in their degree of effectiveness or optimality. However, ``visual search'' encompasses a wide range of specific tasks, each tapping into a different aspect of behaviour (e.g. feature-based attention, information sampling). The aim of the present study is to investigate the extent to which individual differences are stable across different visual search paradigms. Are observers who use the optimal strategy in the SHLS paradigm also more optimal in the ACVS task? Does it make sense to talk about `super-searchers' who show above average performance in a range of search tasks (analogous to the 'super-recognizers' of the face-recognition literature  \citep{russell2009})? As a secondary question, we will measure the test-retest reliability of the differences found in the SHLS paradigm, and compare it with existing estimates of reliability for ACVS and MCFT.

@@

Sample size was determined in part by a power analysis, and in part due to constraints with counter-balancing. $n = 64$ participants means we should be able to detect correlations with $r > 0.34$ with $\alpha = 0.05$, $\beta = 0.80$ between the different visual search paradigms. The sample is therefore of sufficient size to detect relatively small correlations.

@@

The study consists of three paradigms from the visual search literature in which large individual differences have been  found \citep{nowakowska2017, irons-leber2016, kristjansson2014}. Example stimuli can be seen in Figure \ref{fig:exampleStimuli}. A brief overview of each paradigm is given below, with full details in \textit{supplementary materials}. The three tasks were completed over two sessions, approximately one week apart. The SHLS was run in both sessions allowing us to measure test-retest reliability. The order in which participants completed the tasks was counter-balanced. There are 16 different possible orders of tasks/conditions; four participants completed each order for a total of 64.

@@

Stimuli consisted of arrays of black oriented line segments against a grey background. The target was oriented $45^{\circ}$ clockwise, while the distractor items had a random orientation with a mean of $45^{\circ}$ anti-clockwise. The variance was low ($18^{\circ}$) on one half of the display to create a homogeneous texture, and high ($95^{\circ}$) on the other side to create a heterogeneous texture. This means that when the target is present on the homogeneous half, it can be easily be detected with peripheral vision, but when it is in the heterogeneous half, it is much harder to detect. There were 160 trials in total and homo- and heterogeneous sides of the display were randomly varied from trial to trial. The position of the dominant eye was recorded using a desktop-mounted EyeLink 1000 eye tracker (SR Research, Canada).

@@

In the feature foraging task, search displays contained small red, green, yellow and blue circles. For half of the participants, targets were red and green circles, and for the other half of participants, targets were blue and yellow circles. Participants were asked to collect all of the targets within a trial by using the mouse to click on each target. Clicking on a target caused it to disappear from the display. If the participant clicked erroneously on a non-target, the trial immediately ended and a replacement trial began. In the conjunction foraging task, search displays were composed of both circles and squares. For half of the participants, the shapes were red and green, and for the remaining participants the shapes were blue and yellow. Targets were defined by conjunctions of colour and shape (e.g., red squares and green circles, with red circles and green squares as distractors). The assignment of targets and distractors was assigned at random for each participant. The procedure was otherwise identical to the feature foraging task.

@@

Our results are consistent with the original SHLS study \citep{nowakowska2017}: we find a large range of individual differences in search reaction time and accuracy (see Figure \ref{fig:splithalf_summary}). Furthermore, these differences are stable across the two sessions, with Pearson's $r \in [0.71, 0.89]$ ($95\%$ CI.) for accuracy in finding hard targets. We get similar scores for the correlation in reaction times between sessions \textit{a} and \textit{b} for hard targets, ($r \in [0.54-0.81]$), easy targets ($r \in [0.52-0.80]$) and target absent trials ($r \in [0.66-0.86]$).

@@

We have successfully replicated the previous findings around individual differences in visual search strategy in each of the three tasks. Furthermore, the SHLS task has been shown to have good test-retest reliability, similar to that of the ACVS  and MCFT tasks. Given this, we now investigate the extent to which an individual's performance in one of the tasks tells us about how well they will do in the other two.

@@

There are many reasons why two measurements might be uncorrelated, such as range restriction or measurement noise. The test-retest correlations on each of the individual visual search tasks rule out many of these alternatives, leaving a true absence of shared variance between these tasks as a likely explanation. One might have expected reaction time to be at least modestly correlated from one search task to the next, even in the absence of correlations in strategy, as a general factor like an individual's speed-accuracy trade-off, motivation or conscientiousness might lead to better or worse overall performance across all the tasks. However, even reaction times were not highly correlated between tasks. What we did observe were correlations between our measures of strategy and reaction time within each task, suggesting our strategy metrics are predictive of search performance, and reinforcing the interpretation that an individual who is strategic in one task is not necessarily strategic in the others.

@@

Individual differences pose a challenge for efforts to devise a comprehensive model of visual search. Our understanding of the underlying mechanisms of visual search is based predominantly on experiments that systematically vary details of the search task and measure their effects on performance. This approach has led to important insights, for example, about the kinds of visual features that can be used to guide attention (e.g.  \citep{treisman1980, wolfe2007}; how attentional control settings filter out distractors (e.g.  \citep{folk1992,yantis1999}); and  biases in attention, such as a bias towards unexplored locations (e.g.  \citep{klein2000}). These insights are based on search performance averaged over some number of individual participants. For all three of the experiments included in the current study, however, the average performance would be highly misleading, as it would describe very few of the individuals who completed the tasks. In the original SHLS study  \citep{nowakowska2017}, for example, the original aim of the experiment was to assess whether search behaviour could be better described by an optimal  \citep{najemnik-geisler2008}, versus a stochastic  \citep{clarke2016}, model. Considering only the average search performance, the stochastic model was a good explanation. Underlying that average performance, however, was a very large spectrum of search behaviour, replicated here, some of which would be clearly categorized as optimal, and some as stochastic, and some as neither. The original question needed to be refined into new question: for whom is search optimal and for whom is it stochastic? The current findings add even further challenges for researchers interested in understanding visual search, by suggesting we need to not only account for individual differences, but also the interaction of a given individual with a particular search context.

We view these findings, not as a disappointing null result, but as thought-provoking and exciting. Vogel and Awh \citep{vogel2008} argued that studying individual differences in cognitive psychology (in their case, working memory) provides valuable insight to containing potential theories of the underlying cognitive mechanisms. Our results suggest that context and structure of the task also needs to be taken into account and we hope that studying how an individual's behaviour varies across different search tasks will lead to the development of a more comprehensive theory of search.

>>>>

As is common in cognitive psychology, most visual search literature has focused on how the average participant performs in the task, despite it being well known that there is a great deal of variability between one subject and the next. From Treisman's work on Feature Integration Theory \citep{treisman1980} to the latest incarnation of the Guided Search Model \citep{wolfe2015}, we have a good understanding of what makes particular objects easier or harder to find. However, these theories and models have neglected the question of why some observers find visual search so much harder than others. These differences can emerge from several different sources of variation: tiredness \citep{mackworth1948}, information-processing ability, speed-accuracy trade-off, motivation, visual impairments \citep{nowakowska2016}, and search strategies \citep{boot2006}. Although their existence has previously been noted \citep{mackworth1948}, a rigorous examination of individual differences in visual search is a challenge that has not been taken up by many researchers, and questions about their impact and stability remain relatively under-explored.

Here we focus on one source of individual differences in visual search: strategy. By strategy we mean a collection of search behaviours from which all observers can freely choose. Examples include adopting a systematic left-to-right and top-to-bottom strategy \citep{gilchrist2006}, or prioritizing locations that, based on knowledge or context, are more likely to contain the target \citep{wolfe2015}. A striking example of the effect of strategy is given by \cite{boot2006}. They asked participants to monitor a cluttered display for an object changing colour or suddenly appearing. Large individual differences were found with respect to the number of saccades participants made while monitoring the stimulus, which was negatively correlated with detection performance. In another example, Proulx \citeyearpar{proulx2011} demonstrated that participants differed in how they prioritised visual features (colour or orientation) when searching through a set of objects for one with a particular colour-orientation combination. These differences in search strategy explained some of the variation in the extent to which participants were distracted by an irrelevant object of a unique size. However, participants were largely unaware of the search strategy they employed.

Eye movement strategies have also been shown to be an important source of individual differences in visual search efficiency. \cite{nowakowska2017} designed a simple search paradigm to discriminate between optimal \citep{najemnik-geisler2008} and stochastic  \citep{clarke2016} search strategies. Participants searched through arrays of line segments (Figure \ref{fig:exampleStimuli}) arranged such that those on one side of the display all had a very similar orientation (homogeneous), while those on the other side had higher variance (heterogeneous). This meant that targets appearing on the homogeneous side were highly salient, while targets on the heterogeneous side were harder to find. The optimal eye movement strategy here is to search only the heterogeneous half, as targets on the homogeneous side can be detected with peripheral vision. We will refer to this paradigm here as the Split-Half Line Segment task (SHLS). Some participants searched the displays near optimally, but others carried out strategies counter to this, failing to even match the performance of the stochastic searcher. The degree to which participants made saccades in line with the optimal search strategy was strongly correlated with the speed of their search. A related version of this paradigm has been used in research investigating eye movement strategies in response to (simulated) hemianopia \citep{nowakowska2016,nowakowska2018}, with similar conclusions: the full spectrum of individual differences in strategy was observed, from near-optimal to the opposite of optimal. It is therefore not possible to conclude whether optimal or stochastic models better describe search without first explaining individual variability.

@@

Another example of differences in search strategy comes from the foraging literature  \citep{kristjansson2014,johannesson2016}. In this context, foraging tasks involve searching for multiple targets on each trial. Participants were asked to search through a set of items from four categories, with two categories classed as targets. In the conjunction condition (searching for red-horizontal and green-vertical line segments among red-vertical and green-horizontal distracters), most observers searched in runs, finding all the targets of one target category, and then switching and finding the targets in the other category. This strategy has previously been observed in animal foraging  \citep{dawkins1971}, and suggests holding one complex target template in mind at a time is a better strategy than switching templates. However, a sub-set of participants, termed `super-foragers', were able to change between search target categories with no cost to performance. This study has been run on touch displays \citep{kristjansson2014} (observers are asked to tap on as many targets as they can as quickly as possible) and a 3D computer game (in which participants navigate a squirrel through a virtual environment picking up targets with different feature combinations) \citep{prpic2018}, with super-foragers present in both versions. While test-retest reliability has not been measured explicitly for the foraging paradigm, the task was used as a measure to assess the effect of a six day long mindfulness retreat on cognitive performance \citep{hartkamp2017}. From a re-analysis of these data, we can estimate that the test-retest reliability for the mean run length is $r \approx 0.7$ for the feature condition and $r \approx 0.88$ for the conjunction search.

@@

A common theme emerging from these studies is the observation that individual strategies vary in their degree of effectiveness or optimality. However, ``visual search'' encompasses a wide range of specific tasks, each tapping into a different aspect of behaviour (e.g. feature-based attention, information sampling). The aim of the present study is to investigate the extent to which individual differences are stable across different visual search paradigms. Are observers who use the optimal strategy in the SHLS paradigm also more optimal in the ACVS task? Does it make sense to talk about `super-searchers' who show above-average performance in a range of search tasks (analogous to the 'super-recognizers' of the face-recognition literature  \citep{russell2009})? As a secondary question, we will measure the test-retest reliability of the differences found in the SHLS paradigm, and compare it with existing estimates of reliability for ACVS and MCFT.

@@

Sample size was determined in part by a power analysis, and in part by counter-balancing. With $n = 64$ participants, correlations with $r > 0.34$ with $\alpha = 0.05$, $\beta = 0.80$ between the different visual search paradigms can be detected. The sample is therefore of sufficient size to detect relatively small correlations.

@@

The study consists of three paradigms from the visual search literature in which large individual differences have been  found \citep{nowakowska2017, irons-leber2016, kristjansson2014}. Example stimuli can be seen in Figure \ref{fig:exampleStimuli}. A brief overview of each paradigm is given below, with full details in \textit{supplementary materials}. The three tasks were completed over two sessions, approximately one week apart. The SHLS was run in both sessions. The order in which participants completed the tasks was counter-balanced. There are 16 different possible orders of tasks/conditions; four participants completed each order for a total of 64.

@@

Stimuli consisted of arrays of black oriented line segments against a grey background. The target was oriented $45^{\circ}$ clockwise, while the distractor items had a random orientation with a mean of $45^{\circ}$ anti-clockwise. The variance was low ($18^{\circ}$) on one half of the display to create a homogeneous texture, and high ($95^{\circ}$) on the other side to create a heterogeneous texture. When the target is present on the homogeneous half, it can be easily be detected with peripheral vision, but when it is in the heterogeneous half, it is much harder to detect (this was verified in \cite{nowakowska2017}). There were 160 trials in total and homo- and heterogeneous sides of the display were randomly varied from trial to trial. The dominant eye position was recorded using a desktop-mounted EyeLink 1000 eye tracker (SR Research, Canada).

@@

In the feature foraging condition, search displays contained small red, green, yellow and blue circles. For half of the participants, targets were red and green circles, and for the other half of participants, targets were blue and yellow circles. Participants were asked to collect all of the targets within a trial by using the mouse to click on each target. Clicking on a target caused it to disappear from the display. If the participant clicked erroneously on a non-target, the trial immediately ended and a replacement trial began. The conjunction foraging task was the same, except search displays were composed of both circles and squares. For half of the participants, the shapes were red and green, and for the remaining participants the shapes were blue and yellow. Targets were defined by conjunctions of colour and shape (e.g., red squares and green circles, with red circles and green squares as distractors). The assignment of targets and distractors was assigned at random for each participant. The procedure was otherwise identical to the feature foraging task.

@@

Our results are consistent with the original SHLS study \citep{nowakowska2017}: we find a large range of individual differences in search reaction time and accuracy (see Figure \ref{fig:splithalf_summary}). These differences are stable across the two sessions, with Pearson's $r \in [0.71, 0.89]$ ($95\%$ CI.) for accuracy in finding hard targets. We get similar scores for the correlation in reaction times between sessions \textit{a} and \textit{b} for hard targets, ($r \in [0.54-0.81]$), easy targets ($r \in [0.52-0.80]$) and target absent trials ($r \in [0.66-0.86]$).

@@

We have successfully replicated the previous findings around individual differences in visual search strategy in each of the three tasks. Furthermore, the SHLS task has been shown to have good test-retest reliability, similar to that of the ACVS  and MCFT tasks. Given this, we can report the extent to which an individual's performance in one of the tasks predicts performance in the other two.

@@

There are many reasons why two measurements might be uncorrelated, such as range restriction or measurement noise. The test-retest correlations on each of the individual visual search tasks rule out many of these alternatives, leaving a true absence of shared variance between these tasks as a likely explanation. One might have expected reaction time to be at least modestly correlated from one search task to the next, even in the absence of correlations in strategy, as a general factor like an individual's speed-accuracy trade-off, motivation or conscientiousness might lead to better or worse overall performance across all the tasks. However, what we did observe were high correlations between our measures of strategy and reaction time within each task, suggesting our strategy metrics determine a large proportion of search performance, and reinforcing the interpretation that an individual who is strategic in one task is not necessarily strategic in the others.

@@

Individual differences pose a challenge for efforts to devise a comprehensive model of visual search. Our understanding of the mechanisms of visual search is based predominantly on experiments that systematically vary details of the search task and measure their effects on averaged performance. This approach has led to important insights, for example, about the kinds of visual features that can guide attention (e.g.  \citep{treisman1980, wolfe2007}; how attentional control settings filter  distractors (e.g.  \citep{folk1992,yantis1999}); and biases in attention, such as a bias towards unexplored locations (e.g.  \citep{klein2000}). For all three of the experiments included in the current study, however, the average performance would be highly misleading, as it would describe very few of the individuals' performance. In the original SHLS study  \citep{nowakowska2017}, for example, the original aim of the experiment was to assess whether search behaviour could be better described by an optimal  \citep{najemnik-geisler2008}, versus a stochastic  \citep{clarke2016}, model. Considering only the average search performance, the stochastic model was a good explanation. Underlying that average performance, however, was a large spectrum of search behaviour, replicated here, some of which would be clearly categorized as optimal, and some as stochastic, and some as neither. The original question needed to be refined into new question: for whom is search optimal and for whom is it stochastic? The current findings add even further challenges for researchers interested in understanding visual search, by suggesting we need to account not only for individual differences, but also for the interaction of a given individual with a particular search context.

We view these findings not as a discouraging result, but as thought-provoking and exciting. Vogel and Awh \citep{vogel2008} argued that studying individual differences in cognitive psychology (in their case, working memory) provides valuable insight to containing potential theories of the underlying cognitive mechanisms. Our results suggest that context and structure of the task also needs to be taken into account. Understanding how an individual's behaviour varies across different search tasks can lead to the development of a more comprehensive theory of search.
